### 2014-03-25 Pero_Transcriptome.tex
Tuesday PM

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

@@

For biologists interested in understanding the relationship between fitness, genotype, and phenotype, modern sequencing technologies provide for an unprecedented opportunity to gain a deep understanding of genome level processes that together, underlie adaptation.  One interesting example of adaptation lies in animals ability to survive desert conditions. Here, heat \emph{and} drought provide for powerful selective forces, testing animals' ability to osmoregulate and thus to survive. \\

Specifically, the maintenance of water balance in animals is one of the most important physiologic processes, and is critical to desert survival. Indeed, mammals are exquisitely sensitive to changes in osmolality, with slight derangement eliciting physiologic compromise.  When the loss of water exceeds dietary intake, dehydration - and in extreme cases, death - can occur.  Unlike most mammals, animals living in desert habitats are subjected to long periods of extreme heat and intense drought.  As a result, desert animals have evolved mechanisms through which physiologic homeostasis is maintained despite severe and prolonged dehydration. \\

One such desert-adapted rodent, a cricetid rodent endemic to the Southwest United states is a novel model for the study of adaptation to desert environments. They have a lifespan typical of small mammals, and therefore an individual may live it's entire life without ever drinking water. These rodents have distinct advantage over other desert animals (e.g. \textit{Dipodomys}) in that they breed readily in captivity, which enables laboratory studies of the phenotype of interest. In addition the focal species is positioned in a clade of well known animals (e.g. \textit{P. californicus}, \textit{P. maniculatus} and \textit{P. polionotus}) \cite{Bradley:2007jo}. There are growing genetic and genomic resources available \cite{Shorter:2014kw,Panhuis:2011co,Shorter:2012ib}. \\

@@

To begin to understand how genes may underlie desert adaptation, I collected ~15 individuals from a single population \textit{P. eremicus} over a two year time period (2012-2013). These individuals were captured in live traps, then euthanized using isoflurane overdose and decapitation. Immediately post-mortem, the abdominal and pelvis organs were removed, cut in half (in the case of kidney), placed in RNAlater and flash frozen in liquid Nitrogen. Removal of the brain, with similar preservation techniques followed that. Time from euthanasia to removal of all organs never exceeded five minutes. Samples were transferred to a -80C freezer at a later date. These procedures were approved by the University of California Berkeley Animal Care and Use Committee and follow guidelines established by the American Society of Mammalogy for the use of wild animals in research \cite{Sikes:2011dz}.

@@

Following recommendations from MacManes \cite{MacManes:2014io} and Mbandi \cite{Christoffels:2014bg}, adapter sequence contamination was removed, and low quality nucleotides (defined as \textsc{Phred} \textless 2) were removed from the dataset using the program Trimmomatic version 0.32  \cite{Lohse:2012fg}. We concatenated sequence data from each reference tissue type and assembled them using the Trinity beta version released 16 March 2014 \cite{Haas:2013jq}. We used flags indicating the stranded nature of sequencing reads and set maximum allowable physical distance between read pairs to 999nt. The assembly was conducted on the \textsc{Xsede} computer resource Blacklight. To filter the raw sequence assembly, I estimated TPM for each assembled sequence using bwa-mem version 0.77 \cite{Li:2013wn} and eXpress version 1.51 \cite{Roberts:2012dh}, removing all contigs whose expression was less than TPM=1 \cite{Haas:2013jq}. \\

>>>>

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm
\linespread{1.2}
% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

@@

For biologists interested in understanding the relationship between fitness, genotype, and phenotype, modern sequencing technologies provide for an unprecedented opportunity to gain a deep understanding of genome level processes that together, underlie adaptation. Unlike more traditional approaches

 One interesting example of adaptation lies in animals ability to survive desert conditions. Here, heat \emph{and} drought provide for powerful selective forces, testing animals' ability to osmoregulate and thus to survive. \\

Specifically, the maintenance of water balance in animals is one of the most important physiologic processes, and is critical to desert survival \cite{Walsberg:2000uu}. Indeed, mammals are exquisitely sensitive to changes in osmolality, with slight derangement eliciting physiologic compromise.  When the loss of water exceeds dietary intake, dehydration - and in extreme cases, death - can occur.  Unlike most mammals, animals living in desert habitats are subjected to long periods of extreme heat and intense drought.  As a result, desert animals have evolved mechanisms through which physiologic homeostasis is maintained despite severe and prolonged dehydration. \\

One such desert-adapted rodent, a cricetid rodent endemic to the Southwest United States \cite{Veal:2001vp}, is a novel model for the study of adaptation to desert environments. They have a lifespan typical of small mammals, and therefore an individual may live it's entire life without ever drinking water. These rodents have distinct advantage over other desert animals (e.g. \textit{Dipodomys}) in that they breed readily in captivity, which enables laboratory studies of the phenotype of interest. In addition the focal species is positioned in a clade of well known animals (e.g. \textit{P. californicus}, \textit{P. maniculatus} and \textit{P. polionotus}) \cite{Bradley:2007jo}. There are growing genetic and genomic resources available \cite{Shorter:2014kw,Panhuis:2011co,Shorter:2012ib}. \\

@@

To begin to understand how genes may underlie desert adaptation, I collected ~15 individuals from a single population \textit{P. eremicus} over a two year time period (2012-2013). These individuals were captured in live traps, then euthanized using isoflurane overdose and decapitation. Immediately post-mortem, the abdominal and pelvic organs were removed, cut in half (in the case of kidney), placed in RNAlater and flash frozen in liquid Nitrogen. Removal of the brain, with similar preservation techniques, followed that. Time from euthanasia to removal of all organs never exceeded five minutes. Samples were transferred to a -80C freezer at a later date. These procedures were approved by the University of California Berkeley Animal Care and Use Committee and follow guidelines established by the American Society of Mammalogy for the use of wild animals in research \cite{Sikes:2011dz}.

@@

Following recommendations from MacManes \cite{MacManes:2014io} and Mbandi \cite{Christoffels:2014bg}, adapter sequence contamination was removed, and low quality nucleotides (defined as \textsc{Phred} \textless 2) were removed from the dataset using the program Trimmomatic version 0.32  \cite{Lohse:2012fg}. We concatenated sequence data from each reference tissue type and assembled them jointly using the Trinity beta version released 16 March 2014 \cite{Haas:2013jq}. We used flags indicating the stranded nature of sequencing reads and set maximum allowable physical distance between read pairs to 999nt. The assembly was conducted on the \textsc{Xsede} computer resource Blacklight. To filter the raw sequence assembly, I estimated TPM for each assembled sequence using bwa-mem version 0.75 \cite{Li:2013wn} and eXpress version 1.51 \cite{Roberts:2012dh}, removing all contigs whose expression was less than TPM=1 \cite{Haas:2013jq}. \\
